Cycle Route Uptake and Scenario Estimation (CRUSE): an open and nationally scalable approach for strategic cycle network planning tool development \\\large{A case study of Ireland}

### 2024-03-22 paper.qmd
Updates

```{r}
#| include: false
#| eval: false
library(tidyverse)
remotes::install_dev("tmap")
library(tmap)
# Save title page (1st page of paper.pdf):
system("pdftk _site/paper.pdf cat 1 output _site/title_page.pdf")
# Combine cv.pdf with visions.pdf:
system("pdftk _site/cv.pdf _site/visions.pdf cat output _site/cv_and_visions.pdf")
# latexdiffr::latexdiff("paper-v1.qmd", "paper.qmd")
system("latexdiff paper-v1.tex paper.tex > paper-diff.tex")
system("pdflatex paper-diff.tex")
```

>>>>

```{r}
#| include: false
#| eval: false
library(tidyverse)
remotes::install_dev("tmap")
library(tmap)
# Save title page (1st page of paper.pdf):
system("pdftk _site/paper.pdf cat 1 output _site/title_page.pdf")
# Combine cv.pdf with visions.pdf:
system("pdftk _site/cv.pdf _site/visions.pdf cat output _site/cv_and_visions.pdf")
# latexdiffr::latexdiff("paper-v1.qmd", "paper.qmd")
system("latexdiff paper-v1.tex paper.tex > paper-diff.tex")
system("pdflatex paper-diff.tex")
browseURL("paper-diff.pdf")
```

### 2024-03-22 paper.tex
Updates

Transport systems dominated by heavy and powerful private cars are
inefficient, dangerous, and unhealthy. Cars are a major and growing
source of greenhouse gas emissions causing climate change
\citep{winkler2023}, a leading cause of premature death and injury due
to road traffic collisions \citep{globals2018} , and a cause of disease
due to physical inactivity and air, noise and (from tyres) microplastic
pollution \citep{mattsson2023} \citep{welch2023} \citep{cavallaro2024}.

Transport is responsible for 23\% of global greenhouse gas (GHG)
emissions, 70\% of which are from road transport. Nearly half of these
emissions (around 10\% of global emissions) are from passenger cars
\citep{jaramillo2022}. The transport system encourages, enables and in
some cases enforces unsustainable lifestyles. Services that are only
accessible by car lock-in car dependency
\citep{gray2001, shergold2012, motte-baumvol2010}.

Recognizing the growing evidence of such impacts of poorly designed and
performing transport systems, governments in many countries have set
targets and taken actions. In the context of climate, road safety and
physicial inactivity crises, policies to improve transport systems can
be classified according to the `Avoid-Shift-Improve' (ASI) framework
\citep{jaramillo2022}. The ASI framework highlights the importance of
demand reduction (\emph{avoid}ing unnecessary trips), in addition to
mode \emph{shift} to sustainable modes and \emph{improvement} of
existing energy converters, in that order.

@@

Another motivation for cycle network planning is the Road Infrastructure
Safety Management (RISM) directive (2008/96/EC), which requires member
states to implement a road safety management system (RSMS) for all
public roads. Specifically, ``Member States shall ensure that the
ranking of high accident concentration sections and the network safety
ranking are carried out'' \citep{directiv2008}. Given that `safety' in
this context is usefully quantified as the number of people killed and
seriously injured (KSI) per distance traveled, the directive requires
estimation of distance traveled by mode, down to the road link level.
For active modes, about which there is a paucity of data compared with
motorized modes, this is a major challenge. Better data to inform road
safety policies and interventions is a motivation for the CRUSE Tool
outlined in this paper, and modeling active travel more broadly.

In Ireland, the Road Safety Authority (RSA) has set the target of
halving the number of road traffic deaths and serious injuries by 2030
\citep{national2021}. Doing so while simultaneously enabling rapid
uptake of active modes will require key active travel routes to be
identified and improved. Cycling in Ireland represents only 3\% of total
modal share as of the latest Census (2016), but accounts for 20\% of
serious injuries and 7\% of all fatalities. A proactive approach to the
development of cycling policy and infrastructure is taking place in
Ireland. The main frameworks underpinning these efforts are the Climate
Action Plan \citep{climate2022a}, the National Development Plan, and the
National Roads 2040 strategy \citep{national2023}. These policy
frameworks are backed-up by public support for active travel
interventions: while only 3\% of trips in Ireland are made by cycling
according to the latest Census data from 2016 ``83\% of respondents
supported enhanced walking and cycling facilities''
\citep{sustaina2019}.

At the regional level, the recently published Greater Dublin Area
Transport Strategy \citep{greater} reinforces these findings: ``nearly a
quarter of adults cycle at least once a week in the Dublin Metropolitan
Area'' with cycling in the Dublin area taking up to 60,000 cars off the
road today. Extrapolating this on a per population basis across Ireland,
with around 40\% of the population living in Dublin, suggests that
around 150,000 cars could be removed nationwide just by achieving Dublin
levels of cycling in all counties (notwithstanding existing cycling
trips and differences in trip distances). 7 in 10 trips in Ireland are
by car \citep{national} . Cycling has the potential to replace a large
proportion of these trips: ``A high priority must also be given to
cyclists, because trips by this mode have the potential to replace trips
by private car, most specifically for short to medium distance trips,
but increasingly for longer trips as ebikes extend the range of this
mode'' \citep{greater}.

Further evidence of the importance of cycling in Ireland \emph{already}
is provided by the National Strategic Objective (NSO) from the National
Development Plan, with €8.6 billion allocated to sustainable transport
infrastructure including public transport and active travel
interventions. Cycle infrastructure will be developed in synchrony with
the BusConnects project, an entire redesign of the bus network in Dublin
and Cork.

It was in this context that Transport Infrastructure Ireland (TII)
commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool
for Ireland. Based on the Propensity to Cycle Tool (PCT) for England and
Wales, the CRUSE Tool was developed to provide evidence on current
cycling levels and future cycling potential nationwide across Ireland.

A key consideration for TII is to ensure both urban and rural Ireland
were targeted for cycling infrastructure, as shown in
Figure~\ref{fig-dublin}. To this end, TII extensively reviewed activity
on Irish Roads and how the road network was being used by commuters,
cyclists, heavy goods vehicles and other road users across the Irish
transport network. The aim was for the tool to provide strong, national,
systematic but locally-specific evidence to monitor cycling friendliness
and safety in Ireland, to ensure strategic alignment with national,
regional and local policies. The open source and publicly available
nature of the tool ensures more inclusive and evidence-based
conversations around cycle network planning between all stakeholders in
the transport planning process.

TII also requires evidence to support monitoring of growth in cycling,
upgrades of existing cycling infrastructure, business case development
and appraisal, and exposure information to estimate crash rates.
Exposure information is a key part of the the European RISM Directive,
for reporting collision rates of vulnerable road users by 2024. These
factors mean that the tool can be seen as an open access `leverage
point', providing key information for and supporting ambitious plans in
many aspects the planning system, from network design to post-build
monitoring \citep{lovelace2020}.

@@

As the benefits of strategic planning for cycling become more apparent
\citep{scappini2022} we expect the demand for tools such as CRUSE to
grow.

>>>>

Transport systems dominated by heavy and powerful private cars are
inefficient, dangerous, and unhealthy. Cars are a major and growing
source of greenhouse gas emissions causing climate change
\citep{winkler2023}, a leading cause of premature death and injury due
to road traffic collisions \citep{globals2018} , and a cause of disease
due to physical inactivity and air, noise and (from tyres) microplastic
pollution \citep{mattsson2023} \citep{welch2023} \citep{cavallaro2024}.
Transport is responsible for 23\% of global greenhouse gas (GHG)
emissions, 70\% of which are from road transport. Nearly half of these
emissions (around 10\% of global emissions) are from passenger cars
\citep{jaramillo2022}. The transport system encourages, enables and in
some cases enforces unsustainable lifestyles. Services that are only
accessible by car lock-in car dependency
\citep{gray2001, shergold2012, motte-baumvol2010}.

Growing evidence of the negative impacts of car-dependent transport
systems has led governments in many countries to set targets and taken
actions. In the context of climate, road safety and physicial inactivity
crises, policies to improve transport systems can be classified
according to the `Avoid-Shift-Improve' (ASI) framework
\citep{jaramillo2022}. The ASI framework highlights the importance of
demand reduction (\emph{avoid}ing unnecessary trips), in addition to
mode \emph{shift} to sustainable modes and \emph{improvement} of
existing energy converters, in that order.

@@

Another motivation for cycle network planning at the European level is
the Road Infrastructure Safety Management (RISM) directive (2008/96/EC),
which requires member states to implement a road safety management
system (RSMS) for all public roads. Specifically, ``Member States shall
ensure that the ranking of high accident concentration sections and the
network safety ranking are carried out'' \citep{directiv2008}. Given
that `safety' in this context is usefully quantified as the number of
people killed and seriously injured (KSI) per distance traveled, the
directive requires estimation of distance traveled by mode, down to the
road link level. For active modes, about which there is a paucity of
data compared with motorized modes, this is a major challenge. Better
data to inform road safety policies and interventions is a motivation
for better estimates of `baseline' levels of physical activity at high
levels of geographic resolution \citep{tait2023}.

National governments are increasingly acting on the evidence. In
Ireland, the Road Safety Authority (RSA) has set the target of halving
the number of road traffic deaths and serious injuries by 2030
\citep{national2021}. Doing so while simultaneously enabling rapid
uptake of active modes will require key travel corridors to be
identified `cycle proofed'. Cycling in Ireland represents only 3\% of
total modal share as of the 2016 Census, but accounts for 20\% of
serious injuries and 7\% of all fatalities, driving the development of
cycling policy and infrastructure plans in Ireland. The main frameworks
underpinning these efforts are the Climate Action Plan
\citep{climate2022a}, the National Development Plan, and the National
Roads 2040 strategy \citep{national2023}. These policy frameworks are
backed-up by public support for active travel interventions: while only
3\% of trips in Ireland are made by cycling according to the latest
Census data from 2016 ``83\% of respondents supported enhanced walking
and cycling facilities'' \citep{sustaina2019}.

At the regional level within Ireland, the recently published Greater
Dublin Area Transport Strategy \citep{greater} reinforces these findings
of high support for and capability for cycling: ``nearly a quarter of
adults cycle at least once a week in the Dublin Metropolitan Area'' with
cycling in the Dublin area taking up to 60,000 cars off the road today.
Extrapolating this on a per population basis across Ireland, with around
40\% of the population living in Dublin, suggests that around 150,000
cars could be removed nationwide just by achieving Dublin levels of
cycling in all counties (notwithstanding existing cycling trips and
differences in trip distances). 7 in 10 trips in Ireland are by car
\citep{national} . Cycling has the potential to replace a large
proportion of these trips: ``A high priority must also be given to
cyclists, because trips by this mode have the potential to replace trips
by private car, most specifically for short to medium distance trips,
but increasingly for longer trips as ebikes extend the range of this
mode'' \citep{greater}.

Further evidence of the importance of cycling in Ireland is provided by
the National Strategic Objective (NSO) from the National Development
Plan, which allocates €8.6 billion to sustainable transport
infrastructure including public transport and active travel
interventions. Cycle infrastructure will be developed in synchrony with
the BusConnects project, an entire redesign of the bus network in Dublin
and Cork. It was in this context that Transport Infrastructure Ireland
(TII) commissioned the research reported in this paper, which led to the
Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland.
Building on previous work, including the Propensity to Cycle Tool (PCT)
for England and Wales, the CRUSE Tool was developed to provide evidence
on current cycling levels and future cycling potential nationwide across
Ireland.

Stakeholder consultation by TII emphasized the need for a strong,
national, systematic but locally-specific evidence base on cycle
networks in both urban and rural Ireland. The evidence base had to scale
nationally to support strategic alignment with national, regional and
local policies, but also be useful for local network planning. TII also
requires

@@

As the benefits of strategic planning for cycling become more apparent
\citep{scappini2022} we expect the demand for tools such as CRUSE to
grow. However, the approach is not without limitations and does not meet
all requirements in Ireland or indeed any country where the methods are
applied.

A limitation of the approach as implemented in the CRUSE tool is that
the results are `static', reducing its ability to support monitoring of
growth in cycling, upgrades of existing cycling infrastructure and
dynamic (year-by-year) exposure information to estimate crash rates.
Exposure information is a key part of the the European RISM Directive,
for reporting collision rates of vulnerable road users by 2024. These
factors mean that the tool can be seen as an open access `leverage
point', providing key information for and supporting ambitious plans in
many aspects the planning system, from network design to post-build
monitoring \citep{lovelace2020}.
